---
title: "Global Life Expectancy"
author: "Parimah Alsafi"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  dpi        = 120,
  cache      = FALSE,
  comment    = "#>"
)
options(scipen = 999, digits = 4)
set.seed(2024)
```

# Setup

## Required packages

```{r pkg-install, eval=FALSE}
# Run once. Nothing below installs packages silently.
core <- c("tidyverse", "knitr", "broom", "patchwork", "scales")
stat <- c("car", "lmtest", "MASS", "sandwich")
mixd <- c("lme4", "lmerTest", "broom.mixed", "performance")
regu <- c("glmnet")
ml   <- c("ranger", "xgboost")
miss <- c("mice")
clsf <- c("nnet", "class")
extra <- c("kableExtra", "GGally", "naniar")   # optional, document degrades gracefully

install.packages(c(core, stat, mixd, regu, ml, miss, clsf, extra))
```

```{r}
remove.packages("stringi")
remove.packages("stringr")

install.packages("stringi", type = "source")
install.packages("stringr")
```

```{r}
library(stringi)
library(stringr)
```


```{r pkg-load}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(readr)
  library(purrr); library(tibble); library(forcats)
  library(knitr)
})

## Optional packages are guarded so the document always knits.
has <- function(p) requireNamespace(p, quietly = TRUE)

PKG <- list(
  car        = has("car"),        lmtest   = has("lmtest"),
  MASS       = has("MASS"),       lme4     = has("lme4"),
  lmerTest   = has("lmerTest"),   glmnet   = has("glmnet"),
  ranger     = has("ranger"),     xgboost  = has("xgboost"),
  mice       = has("mice"),       nnet     = has("nnet"),
  class      = has("class"),      broom    = has("broom"),
  broom.mixed = has("broom.mixed"), patchwork = has("patchwork")
)

tibble(package = names(PKG), available = unlist(PKG)) |>
  arrange(available, package) |>
  kable(caption = "Availability of analysis packages")
```

```{r theme}
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(colour = "grey35", size = 10),
      panel.grid.minor = element_blank(),
      strip.text    = element_text(face = "bold")
    )
)
pal_status <- c("Developed" = "#1b6ca8", "Developing" = "#e07b39")
```

## Analysis configuration

```{r config}
CFG <- list(
  data_path        = "data/raw/Life Expectancy Data.csv",
  drop_missing_y   = TRUE,   
  fix_bmi          = TRUE,    
  fix_population   = FALSE,   
  primary_data     = "imputed",  
  n_folds          = 5,       
  temporal_cutoff  = 2012,    
  test_interpolate = TRUE,  
  run_mice         = TRUE,    
  mice_m           = 5,
  mice_maxit       = 5,
  run_xgboost      = TRUE,
  tune_ml          = TRUE,    
  rf_trees         = 500,
  seed             = 2024
)
set.seed(CFG$seed)
```

---

# Part 1 --- Data Audit

```{r load-data}
normalize_names <- function(x) {
  x <- tolower(trimws(x))
  x <- gsub("[^a-z0-9]+", "_", x)
  x <- gsub("_+", "_", x)
  gsub("^_|_$", "", x)
}

read_life_expectancy <- function(path) {
  candidates <- c(
    path,
    "Life Expectancy Data.csv",
    "data/Life Expectancy Data.csv",
    "data/raw/Life_Expectancy_Data.csv"
  )
  hit <- candidates[file.exists(candidates)]
  if (!length(hit)) {
    stop(
      "Data file not found. Download the Kaggle dataset and place ",
      "'Life Expectancy Data.csv' in data/raw/. Looked in:\n  ",
      paste(candidates, collapse = "\n  "),
      call. = FALSE
    )
  }
  readr::read_csv(hit[1], show_col_types = FALSE, progress = FALSE)
}

raw <- read_life_expectancy(CFG$data_path)
names(raw) <- normalize_names(names(raw))

# regex -> canonical name
rename_map <- c(
  country                = "^country$",
  year                   = "^year$",
  status                 = "^status$",
  life_expectancy        = "life.*expect",
  adult_mortality        = "adult.*mortal",
  infant_deaths          = "infant.*death",
  alcohol                = "^alcohol",
  percentage_expenditure = "percentage.*expend",
  hepatitis_b            = "hepatitis",
  measles                = "measles",
  bmi                    = "^bmi$",
  under_five_deaths      = "under.*five.*death",
  polio                  = "^polio",
  total_expenditure      = "total.*expend",
  diphtheria             = "diphther",
  hiv_aids               = "hiv",
  gdp                    = "^gdp$",
  population             = "^population$",
  thinness_10_19         = "thinness.*(1_19|10_19)",
  thinness_5_9           = "thinness.*5_9",
  income_composition     = "income.*composition",
  schooling              = "schooling"
)

apply_rename <- function(df, map) {
  nm <- names(df)
  for (target in names(map)) {
    idx <- grep(map[[target]], nm)
    if (length(idx) >= 1) nm[idx[1]] <- target
  }
  names(df) <- nm
  df
}

life <- apply_rename(raw, rename_map)

missing_core <- setdiff(c("country", "year", "life_expectancy"), names(life))
if (length(missing_core)) {
  stop("Essential columns not recognised: ", paste(missing_core, collapse = ", "))
}

life <- life |>
  mutate(
    country = as.character(country),
    year    = as.integer(year),
    status  = factor(status, levels = c("Developing", "Developed"))
  )

cat("Dimensions:", nrow(life), "rows x", ncol(life), "columns\n")
```

## Structure of the panel

```{r audit-structure}
audit_overview <- tibble(
  metric = c("Observations", "Countries", "Years", "First year", "Last year",
             "Variables", "Duplicated (country, year) rows"),
  value  = c(nrow(life),
             n_distinct(life$country),
             n_distinct(life$year),
             min(life$year, na.rm = TRUE),
             max(life$year, na.rm = TRUE),
             ncol(life),
             sum(duplicated(life[c("country", "year")])))
)
kable(audit_overview, caption = "Panel overview")
```

```{r audit-types}
var_type <- function(x) {
  if (is.factor(x) || is.character(x)) "categorical"
  else if (is.integer(x)) "integer / temporal"
  else "continuous"
}

audit_vars <- tibble(
  variable  = names(life),
  type      = vapply(life, var_type, character(1)),
  n_missing = vapply(life, function(x) sum(is.na(x)), integer(1)),
  pct_missing = round(100 * n_missing / nrow(life), 1),
  n_unique  = vapply(life, function(x) dplyr::n_distinct(x), integer(1)),
  min       = vapply(life, function(x) if (is.numeric(x)) min(x, na.rm = TRUE) else NA_real_, numeric(1)),
  max       = vapply(life, function(x) if (is.numeric(x)) max(x, na.rm = TRUE) else NA_real_, numeric(1))
)
kable(audit_vars, caption = "Variable dictionary, ranges and missingness", digits = 2)
```

## Observations per country


```{r audit-balance}
per_country <- life |> count(country, name = "n_years")

kable(
  per_country |> count(n_years, name = "n_countries") |> arrange(desc(n_years)),
  caption = "Distribution of the number of observed years per country"
)

per_country |>
  filter(n_years < max(n_years)) |>
  arrange(n_years) |>
  head(15) |>
  kable(caption = "Countries with an incomplete time series (first 15)")
```

```{r audit-balance-plot, fig.height=3.4}
ggplot(per_country, aes(n_years)) +
  geom_bar(fill = "#1b6ca8", alpha = .85) +
  labs(title = "Panel is close to balanced but not exactly balanced",
       subtitle = "Number of observed years per country",
       x = "Years observed", y = "Countries")
```

```{r audit-implausible}
rule <- function(data, var, condition, description) {
  if (!var %in% names(data)) return(NULL)
  n <- sum(condition, na.rm = TRUE)
  tibble(variable = var, rule = description, n_flagged = n,
         pct = round(100 * n / nrow(data), 2))
}

flags <- bind_rows(
  rule(life, "life_expectancy", life$life_expectancy < 20 | life$life_expectancy > 100,
       "outside 20-100 years"),
  rule(life, "bmi", life$bmi < 10 | life$bmi > 60,
       "national mean BMI outside 10-60"),
  rule(life, "adult_mortality", life$adult_mortality < 20,
       "adult mortality < 20 per 1000 (suspected truncated digit)"),
  rule(life, "infant_deaths", life$infant_deaths == 0,
       "exactly zero infant deaths"),
  rule(life, "under_five_deaths", life$under_five_deaths == 0,
       "exactly zero under-five deaths"),
  rule(life, "population", life$population < 1000,
       "population below 1000"),
  rule(life, "percentage_expenditure", life$percentage_expenditure == 0,
       "zero health expenditure share")
)
kable(flags, caption = "Range / plausibility checks on the raw data")
```

```{r audit-fix}
n_bmi <- sum(!is.na(life$bmi) & (life$bmi < 10 | life$bmi > 60))
if (CFG$fix_bmi) {
  life <- life |> mutate(bmi = ifelse(!is.na(bmi) & (bmi < 10 | bmi > 60), NA_real_, bmi))
}
if (CFG$fix_population) {
  life <- life |> mutate(population = ifelse(!is.na(population) & population < 1000,
                                             NA_real_, population))
}
cat(sprintf("BMI values sent to NA: %s\nPopulation values sent to NA: %s\n",
            if (CFG$fix_bmi) n_bmi else "0 (rule disabled - flagged only)",
            if (CFG$fix_population) "applied" else "0 (rule disabled - flagged only)"))
```



```{r derive}
log1p_safe <- function(x) ifelse(is.na(x), NA_real_, log1p(pmax(x, 0)))

life <- life |>
  mutate(
    log_gdp        = log1p_safe(gdp),
    log_population = log1p_safe(population),
    log_measles    = log1p_safe(measles),
    log_hiv        = log1p_safe(hiv_aids),
    year_c         = year - min(year, na.rm = TRUE)   # 0 = first observed year
  )

if (CFG$drop_missing_y) {
  n_before <- nrow(life)
  life <- life |> filter(!is.na(life_expectancy))
  cat("Rows dropped for a missing outcome:", n_before - nrow(life), "\n")
}

life <- life |> arrange(country, year)
```

---

# Part 2 --- Missing Data Analysis


## Missing rate per variable


```{r miss-rate, fig.height=5.5}
miss_tbl <- life |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  mutate(missing_rate = n_missing / nrow(life)) |>
  arrange(desc(missing_rate))

kable(miss_tbl |> filter(n_missing > 0), digits = 3,
      caption = "Variables with missing values")

miss_tbl |>
  filter(n_missing > 0) |>
  ggplot(aes(x = missing_rate, y = fct_reorder(variable, missing_rate))) +
  geom_col(fill = "#c0392b", alpha = .85) +
  geom_text(aes(label = scales::percent(missing_rate, accuracy = .1)),
            hjust = -0.12, size = 3.2) +
  scale_x_continuous(labels = scales::percent, expand = expansion(mult = c(0, .18))) +
  labs(title = "Missing rate by variable", x = "Missing rate", y = NULL)
```

## Missingness pattern

```{r miss-pattern}
miss_vars <- miss_tbl |> filter(n_missing > 0) |> pull(variable)

pattern <- life |>
  select(all_of(miss_vars)) |>
  mutate(across(everything(), ~ as.integer(is.na(.x)))) |>
  group_by(across(everything())) |>
  tally(name = "n_rows") |>
  ungroup() |>
  arrange(desc(n_rows))

cat("Rows that are fully complete:",
    sum(complete.cases(life)), "of", nrow(life),
    sprintf("(%.1f%%)\n", 100 * sum(complete.cases(life)) / nrow(life)))

head(pattern, 8) |>
  kable(caption = "Eight most frequent missingness patterns (1 = missing)")
```

```{r miss-by-country}
miss_by_country <- life |>
  select(country, all_of(miss_vars)) |>
  pivot_longer(-country, names_to = "variable", values_to = "value",
               values_transform = list(value = as.numeric)) |>
  group_by(country) |>
  summarise(pct_missing = 100 * mean(is.na(value)), .groups = "drop") |>
  arrange(desc(pct_missing))

head(miss_by_country, 12) |>
  kable(digits = 1, caption = "Countries with the highest share of missing cells")
```

```{r miss-heat, fig.height=6.5}
life |>
  select(country, year, all_of(miss_vars)) |>
  pivot_longer(-c(country, year), names_to = "variable", values_to = "value") |>
  group_by(variable, year) |>
  summarise(missing_rate = mean(is.na(value)), .groups = "drop") |>
  ggplot(aes(year, fct_rev(variable), fill = missing_rate)) +
  geom_tile(colour = "white", linewidth = .3) +
  scale_fill_gradient(low = "#f7f7f7", high = "#c0392b", labels = scales::percent) +
  labs(title = "Missingness is structured in time, not random",
       subtitle = "Share of countries missing each variable, by year",
       x = NULL, y = NULL, fill = "Missing")
```



```{r impute-strategies}
NUM_VARS <- life |>
  select(where(is.numeric)) |>
  select(-any_of(c("year", "year_c"))) |>
  names()

DERIVED_VARS <- c("log_gdp", "log_population", "log_measles", "log_hiv")
# never impute the outcome; derived logs are rebuilt after imputation
IMPUTE_VARS  <- setdiff(NUM_VARS, c("life_expectancy", DERIVED_VARS))

## --- Strategy 1: complete cases -------------------------------------------
dat_cc <- life |> tidyr::drop_na(all_of(c("life_expectancy", IMPUTE_VARS)))

## --- Strategy 2: panel-aware imputation, as an estimator -------------------

# interior gaps only - no extrapolation past the first / last observed year
interp_interior <- function(x, t) {
  ok <- !is.na(x)
  if (sum(ok) < 2) return(x)
  as.numeric(stats::approx(t[ok], x[ok], xout = t, rule = 1)$y)
}

# last observation carried forward: uses the past only
locf <- function(x) {
  idx <- cummax(seq_along(x) * as.integer(!is.na(x)))
  idx[idx == 0] <- NA_integer_
  x[idx]
}

# LEARN the imputation parameters (training data only)
impute_fit <- function(train, vars = IMPUTE_VARS) {
  list(
    vars = vars,
    status_year = train |>
      group_by(status, year) |>
      summarise(across(all_of(vars), ~ stats::median(.x, na.rm = TRUE)), .groups = "drop"),
    global = train |>
      summarise(across(all_of(vars), ~ stats::median(.x, na.rm = TRUE)))
  )
}

impute_apply <- function(newdata, params, history = NULL,
                         mode = c("interior", "causal")) {
  mode <- match.arg(mode)
  vars <- params$vars
  new  <- newdata |> mutate(.rid = dplyr::row_number())
  cols <- c("country", "year", vars, ".rid")

  hist_part <- if (is.null(history)) NULL else
    history |> mutate(.rid = NA_integer_) |> select(all_of(cols))

  comb <- bind_rows(hist_part, new |> select(all_of(cols))) |>
    arrange(country, year) |>
    group_by(country)

  comb <- if (mode == "interior") {
    comb |> mutate(across(all_of(vars), ~ locf(interp_interior(.x, year))))
  } else {
    comb |> mutate(across(all_of(vars), locf))     # backwards-looking only
  }

  filled <- comb |> ungroup() |> filter(!is.na(.rid)) |> arrange(.rid)
  new[vars] <- filled[vars]

  new <- left_join(new, params$status_year, by = c("status", "year"),
                   suffix = c("", ".sy"))
  for (v in vars) {
    sy <- paste0(v, ".sy")
    if (sy %in% names(new)) {
      new[[v]] <- ifelse(is.na(new[[v]]), new[[sy]], new[[v]])
      new[[sy]] <- NULL
    }
  }
  for (v in vars) new[[v]] <- ifelse(is.na(new[[v]]), params$global[[v]], new[[v]])

  new |>
    select(-.rid) |>
    mutate(log_gdp        = log1p_safe(gdp),
           log_population = log1p_safe(population),
           log_measles    = log1p_safe(measles),
           log_hiv        = log1p_safe(hiv_aids))
}


dat_imp <- impute_apply(life, impute_fit(life), mode = "interior")

## --- Strategy 3: multiple imputation (mice) is run in Part 20 --------------

tibble(
  dataset  = c("Complete cases", "Panel-aware imputation", "Original"),
  rows     = c(nrow(dat_cc), nrow(dat_imp), nrow(life)),
  countries = c(n_distinct(dat_cc$country), n_distinct(dat_imp$country),
                n_distinct(life$country)),
  remaining_NA = c(sum(is.na(dat_cc[IMPUTE_VARS])),
                   sum(is.na(dat_imp[IMPUTE_VARS])),
                   sum(is.na(life[IMPUTE_VARS])))
) |>
  kable(caption = "What each missing-data strategy costs")
```

```{r impute-cost, fig.height=3.6}
lost <- setdiff(unique(life$country), unique(dat_cc$country))
cat("Countries lost entirely by complete-case analysis:", length(lost), "\n")
if (length(lost)) cat(paste(utils::head(sort(lost), 20), collapse = ", "), "\n")

bind_rows(
  life   |> transmute(life_expectancy, set = "All rows"),
  dat_cc |> transmute(life_expectancy, set = "Complete cases")
) |>
  ggplot(aes(life_expectancy, fill = set)) +
  geom_density(alpha = .45, colour = NA) +
  scale_fill_manual(values = c("All rows" = "grey40", "Complete cases" = "#1b6ca8")) +
  labs(title = "Complete-case filtering shifts the outcome distribution",
       subtitle = "Evidence that the missingness is not ignorable",
       x = "Life expectancy (years)", y = "Density", fill = NULL)
```


```{r pick-primary}
dat <- if (identical(CFG$primary_data, "complete_case")) dat_cc else dat_imp
cat("Primary analysis dataset:", CFG$primary_data,
    "-", nrow(dat), "rows,", n_distinct(dat$country), "countries\n")


panel <- life
stopifnot(identical(panel$country, dat_imp$country),
          identical(panel$year, dat_imp$year))

PRED_FULL <- intersect(
  c("schooling", "income_composition", "adult_mortality", "infant_deaths",
    "under_five_deaths", "alcohol", "percentage_expenditure", "hepatitis_b",
    "log_measles", "bmi", "polio", "total_expenditure", "diphtheria",
    "log_hiv", "log_gdp", "log_population", "thinness_10_19", "thinness_5_9"),
  names(dat)
)
```

---

# Part 3 --- Exploratory Data Analysis

## Distribution of the outcome

```{r eda-outcome, fig.height=6.5}
p_hist <- ggplot(dat, aes(life_expectancy)) +
  geom_histogram(bins = 40, fill = "#1b6ca8", colour = "white", alpha = .9) +
  labs(title = "Histogram", x = "Life expectancy (years)", y = "Count")

p_dens <- ggplot(dat, aes(life_expectancy)) +
  geom_density(fill = "#1b6ca8", alpha = .35, colour = "#1b6ca8") +
  labs(title = "Density", x = "Life expectancy (years)", y = NULL)

p_box <- ggplot(dat, aes(y = life_expectancy)) +
  geom_boxplot(fill = "#1b6ca8", alpha = .5, width = .4) +
  labs(title = "Boxplot", y = "Life expectancy (years)", x = NULL) +
  theme(axis.text.x = element_blank())

p_qq <- ggplot(dat, aes(sample = life_expectancy)) +
  stat_qq(colour = "#1b6ca8", alpha = .5, size = .8) + stat_qq_line() +
  labs(title = "Normal Q-Q", x = "Theoretical", y = "Sample")

if (PKG$patchwork) {
  (p_hist + p_dens) / (p_box + p_qq)
} else {
  print(p_hist); print(p_dens); print(p_box); print(p_qq)
}
```

```{r eda-summary}
num_summary <- function(x) {
  c(n = sum(!is.na(x)), mean = mean(x, na.rm = TRUE), sd = stats::sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    q25 = stats::quantile(x, .25, na.rm = TRUE, names = FALSE),
    median = stats::median(x, na.rm = TRUE),
    q75 = stats::quantile(x, .75, na.rm = TRUE, names = FALSE),
    max = max(x, na.rm = TRUE),
    skew = mean((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) /
           stats::sd(x, na.rm = TRUE)^3)
}

dat |>
  select(all_of(c("life_expectancy", PRED_FULL))) |>
  map_dfr(num_summary, .id = "variable") |>
  kable(digits = 2, caption = "Summary statistics for the outcome and predictors")
```


## Distributions of the main predictors

```{r eda-predictors, fig.height=8}
dat |>
  select(all_of(PRED_FULL)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(value)) +
  geom_histogram(bins = 35, fill = "#e07b39", colour = "white", alpha = .9) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(title = "Predictor distributions", x = NULL, y = NULL) +
  theme(axis.text = element_text(size = 7))
```

## Country trajectories


```{r eda-traj, fig.height=5}
focus <- intersect(
  c("Germany", "Iran (Islamic Republic of)", "India", "Nigeria", "Japan", "Brazil",
    "South Africa", "Rwanda"),
  unique(dat$country)
)
if (length(focus) < 4) focus <- sample(unique(dat$country), 8)

ggplot(dat, aes(year, life_expectancy, group = country)) +
  geom_line(colour = "grey82", linewidth = .3) +
  geom_line(data = filter(dat, country %in% focus),
            aes(colour = country), linewidth = 1) +
  labs(title = "Countries do not share one trajectory",
       subtitle = "Grey: all countries. Coloured: selected countries",
       x = NULL, y = "Life expectancy (years)", colour = NULL)
```



```{r eda-change, fig.height=4.2}
change <- dat |>
  group_by(country) |>
  filter(year %in% range(year)) |>
  summarise(status = dplyr::first(status),
            start = life_expectancy[which.min(year)],
            end   = life_expectancy[which.max(year)],
            gain  = end - start, .groups = "drop")

r_conv <- stats::cor(change$start, change$gain, use = "complete.obs")
conv_txt <- sprintf("Correlation between starting level and subsequent gain: r = %.2f (%s)",
                    r_conv,
                    if (r_conv < -0.1) "convergence" else if (r_conv > 0.1)
                      "divergence" else "no clear pattern")

ggplot(change, aes(start, gain, colour = status)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = .75, size = 2) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = .6) +
  scale_colour_manual(values = pal_status) +
  labs(title = "Starting level against subsequent change",
       subtitle = conv_txt,
       x = "Life expectancy in the first observed year",
       y = "Change over the observation window (years)", colour = NULL)
```

## Developed vs Developing

```{r eda-status, fig.height=4.2}
p1 <- ggplot(dat, aes(status, life_expectancy, fill = status)) +
  geom_violin(alpha = .35, colour = NA) +
  geom_boxplot(width = .18, outlier.size = .6, alpha = .9) +
  scale_fill_manual(values = pal_status) +
  labs(title = "Distribution by development status", x = NULL,
       y = "Life expectancy (years)") +
  theme(legend.position = "none")

p2 <- dat |>
  group_by(status, year) |>
  summarise(m = mean(life_expectancy), .groups = "drop") |>
  ggplot(aes(year, m, colour = status)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = pal_status) +
  labs(title = "Mean over time", x = NULL, y = NULL, colour = NULL)

if (PKG$patchwork) p1 + p2 else { print(p1); print(p2) }
```

### Descriptive comparison


```{r eda-status-desc}
pooled_sd <- function(x, g) {
  s <- split(x, g); n <- lengths(s); v <- vapply(s, stats::var, numeric(1))
  sqrt(sum((n - 1) * v) / (sum(n) - length(n)))
}
grp_means <- tapply(dat$life_expectancy, dat$status, mean)
cohen_d <- unname((grp_means[["Developed"]] - grp_means[["Developing"]]) /
                    pooled_sd(dat$life_expectancy, dat$status))

# country-level means: one row per country, which at least removes the within-country
# replication (it does not fix the unequal series lengths)
country_level <- dat |>
  group_by(country, status) |>
  summarise(mean_le = mean(life_expectancy), .groups = "drop")

bind_rows(
  dat |> group_by(status) |>
    summarise(unit = "country-year rows", n = n(), mean = mean(life_expectancy),
              median = stats::median(life_expectancy), sd = stats::sd(life_expectancy),
              .groups = "drop"),
  country_level |> group_by(status) |>
    summarise(unit = "country means", n = n(), mean = mean(mean_le),
              median = stats::median(mean_le), sd = stats::sd(mean_le), .groups = "drop")
) |>
  select(unit, status, n, mean, median, sd) |>
  kable(digits = 2,
        caption = "Descriptive comparison at two levels of aggregation (no p-values)")

cat(sprintf("Raw difference in means: %.2f years | Cohen's d (descriptive): %.2f\n",
            grp_means[["Developed"]] - grp_means[["Developing"]], cohen_d))
```

### The same comparison with a model that respects the panel


```{r eda-status-mixed, eval=PKG$lme4}
m_status <- (if (PKG$lmerTest) lmerTest::lmer else lme4::lmer)(
  life_expectancy ~ status + (1 | country), data = dat, REML = TRUE)

st <- as.data.frame(stats::coef(summary(m_status)))
st$term <- rownames(st)
st |> as_tibble() |> select(term, everything()) |>
  kable(digits = 3,
        caption = "Status difference estimated with a country random intercept")

vs <- as.data.frame(lme4::VarCorr(m_status))
cat(sprintf("Between-country SD: %.2f years | residual SD: %.2f years\n",
            sqrt(vs$vcov[vs$grp == "country"]), sqrt(vs$vcov[vs$grp == "Residual"])))
```


## Predictors against the outcome

```{r eda-scatter, fig.height=7}
key <- intersect(c("schooling", "income_composition", "adult_mortality", "log_hiv",
                   "log_gdp", "bmi", "alcohol", "thinness_10_19"), names(dat))

dat |>
  select(life_expectancy, status, all_of(key)) |>
  pivot_longer(-c(life_expectancy, status), names_to = "variable", values_to = "value") |>
  ggplot(aes(value, life_expectancy)) +
  geom_point(aes(colour = status), alpha = .22, size = .7) +
  geom_smooth(method = "loess", se = FALSE, colour = "black", linewidth = .7) +
  scale_colour_manual(values = pal_status) +
  facet_wrap(~ variable, scales = "free_x", ncol = 4) +
  labs(title = "Marginal relationships with life expectancy",
       subtitle = "Black line: LOESS smoother, no model assumed",
       x = NULL, y = "Life expectancy (years)", colour = NULL)
```

---

# Part 4 --- Correlation Analysis

```{r corr, fig.height=7.5}
cor_mat <- dat |>
  select(all_of(c("life_expectancy", PRED_FULL))) |>
  cor(use = "pairwise.complete.obs")

cor_long <- as.data.frame(as.table(cor_mat)) |>
  setNames(c("var1", "var2", "r"))

ord <- stats::hclust(stats::as.dist(1 - abs(cor_mat)))$order
lvl <- colnames(cor_mat)[ord]

cor_long |>
  mutate(var1 = factor(var1, levels = lvl), var2 = factor(var2, levels = lvl)) |>
  ggplot(aes(var1, var2, fill = r)) +
  geom_tile(colour = "white", linewidth = .25) +
  scale_fill_gradient2(low = "#c0392b", mid = "white", high = "#1b6ca8",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation matrix (hierarchically ordered)",
       subtitle = "Linear association only - not causality, not independence",
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8))
```

```{r corr-tables}
cor_long |>
  filter(var1 == "life_expectancy", var2 != "life_expectancy") |>
  arrange(desc(abs(r))) |>
  transmute(predictor = var2, correlation_with_life_expectancy = r) |>
  kable(digits = 3, caption = "Predictors ranked by absolute correlation with the outcome")

cor_long |>
  filter(as.character(var1) < as.character(var2), abs(r) > 0.7) |>
  arrange(desc(abs(r))) |>
  kable(digits = 3, caption = "Predictor pairs with |r| > 0.7 (multicollinearity watchlist)")
```

---

# Part 5 --- Baseline Model


```{r metrics-helpers}
rmse <- function(obs, pred) sqrt(mean((obs - pred)^2, na.rm = TRUE))
mae  <- function(obs, pred) mean(abs(obs - pred), na.rm = TRUE)
r2   <- function(obs, pred) 1 - sum((obs - pred)^2, na.rm = TRUE) /
                                sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)

eval_metrics <- function(obs, pred) {
  tibble(RMSE = rmse(obs, pred), MAE = mae(obs, pred), R2 = r2(obs, pred))
}

# broom-free coefficient table (works even without broom installed)
tidy_fit <- function(model, conf = TRUE) {
  co <- summary(model)$coefficients
  out <- tibble(term = rownames(co), estimate = co[, 1], std.error = co[, 2],
                statistic = co[, 3], p.value = co[, 4])
  if (conf) {
    ci <- suppressMessages(stats::confint(model))
    out$conf.low  <- ci[out$term, 1]
    out$conf.high <- ci[out$term, 2]
  }
  out
}

stars <- function(p) cut(p, c(-Inf, .001, .01, .05, .1, Inf),
                         labels = c("***", "**", "*", ".", ""), right = FALSE)
```

```{r baseline}
m_base <- lm(life_expectancy ~ schooling, data = dat)

tidy_fit(m_base) |>
  mutate(sig = stars(p.value)) |>
  kable(digits = 4, caption = "Baseline model: life expectancy on schooling")

bind_cols(
  tibble(model = "Baseline (schooling only)"),
  eval_metrics(dat$life_expectancy, fitted(m_base)),
  tibble(adj_R2 = summary(m_base)$adj.r.squared,
         AIC = stats::AIC(m_base), BIC = stats::BIC(m_base))
) |>
  kable(digits = 3, caption = "In-sample fit of the baseline model")
```

```{r baseline-plot, fig.height=4.2}
ggplot(dat, aes(schooling, life_expectancy)) +
  geom_point(aes(colour = status), alpha = .25, size = .8) +
  geom_smooth(method = "lm", colour = "black", fill = "grey70") +
  scale_colour_manual(values = pal_status) +
  labs(title = "Baseline fit",
       subtitle = "A single straight line through a curved, clustered cloud",
       x = "Years of schooling", y = "Life expectancy (years)", colour = NULL)
```

---

# Part 6 --- Multiple Linear Regression

```{r mlr}
form_full <- stats::reformulate(c(PRED_FULL, "status", "year_c"),
                                response = "life_expectancy")
m_full <- lm(form_full, data = dat)

tidy_fit(m_full) |>
  mutate(sig = stars(p.value)) |>
  arrange(p.value) |>
  kable(digits = 4, caption = "Full multiple regression (ordered by p-value)")
```

```{r mlr-fit}
bind_rows(
  bind_cols(tibble(model = "Baseline"),
            eval_metrics(dat$life_expectancy, fitted(m_base)),
            tibble(adj_R2 = summary(m_base)$adj.r.squared,
                   AIC = AIC(m_base), BIC = BIC(m_base))),
  bind_cols(tibble(model = "Full MLR"),
            eval_metrics(dat$life_expectancy, fitted(m_full)),
            tibble(adj_R2 = summary(m_full)$adj.r.squared,
                   AIC = AIC(m_full), BIC = BIC(m_full)))
) |>
  kable(digits = 3, caption = "Baseline vs full multiple regression (in-sample)")
```

```{r mlr-coefplot, fig.height=6}
tidy_fit(m_full) |>
  filter(term != "(Intercept)") |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(estimate, term)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey45") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = .25,
                 colour = "grey40") +
  geom_point(aes(colour = p.value < .05), size = 2.4) +
  scale_colour_manual(values = c("TRUE" = "#1b6ca8", "FALSE" = "grey65"),
                      labels = c("TRUE" = "p < .05", "FALSE" = "n.s.")) +
  labs(title = "Adjusted associations with life expectancy",
       subtitle = "Coefficients with 95% confidence intervals; units differ per predictor",
       x = "Estimated coefficient", y = NULL, colour = NULL)
```

---

# Part 7 --- Regression Diagnostics

## Residual analysis

```{r diag-plots, fig.height=8}
op <- graphics::par(mfrow = c(2, 2), mar = c(4.2, 4.2, 2.6, 1.2))
plot(m_full, which = 1:4, pch = 20, col = grDevices::adjustcolor("#1b6ca8", .35))
graphics::par(op)
```

```{r diag-resid-year, fig.height=4}
diag_df <- dat |>
  mutate(.fitted = fitted(m_full), .resid = stats::rstandard(m_full))

ggplot(diag_df, aes(year, .resid)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_jitter(width = .2, alpha = .18, size = .7, colour = "#1b6ca8") +
  geom_smooth(method = "loess", se = TRUE, colour = "#c0392b") +
  labs(title = "Residuals still carry a time signal",
       subtitle = "A pooled OLS model does not account for the global trend",
       x = NULL, y = "Standardised residual")
```

```{r diag-resid-country, fig.height=4}
diag_df |>
  group_by(country) |>
  summarise(mean_resid = mean(.resid), .groups = "drop") |>
  ggplot(aes(mean_resid)) +
  geom_histogram(bins = 40, fill = "#e07b39", colour = "white") +
  geom_vline(xintercept = 0, linetype = 2) +
  labs(title = "Residuals are clustered by country",
       subtitle = "Country means far from zero = systematic, unmodelled country effects",
       x = "Mean standardised residual per country", y = "Countries")
```

## Heteroscedasticity

```{r diag-bp, eval=PKG$lmtest}
bp <- lmtest::bptest(m_full)
tibble(test = "Breusch-Pagan", statistic = unname(bp$statistic),
       df = unname(bp$parameter), p.value = bp$p.value) |>
  kable(digits = 4, caption = "Breusch-Pagan test for heteroscedasticity")
```


```{r diag-robust, eval=PKG$lmtest && has("sandwich")}
V_hc3 <- sandwich::vcovHC(m_full, type = "HC3")
V_cl  <- tryCatch(
  sandwich::vcovCL(m_full, cluster = dat$country, type = "HC3"),
  error = function(e) sandwich::vcovCL(m_full, cluster = ~ country, type = "HC3")
)

rob_hc3 <- lmtest::coeftest(m_full, vcov. = V_hc3)
rob_cl  <- lmtest::coeftest(m_full, vcov. = V_cl)

se_tbl <- tibble(term = rownames(rob_cl),
                 estimate     = rob_cl[, 1],
                 se_classical = summary(m_full)$coefficients[, 2],
                 se_HC3       = rob_hc3[, 2],
                 se_cluster   = rob_cl[, 2],
                 p_cluster    = rob_cl[, 4]) |>
  filter(term != "(Intercept)") |>
  mutate(inflation_vs_classical = se_cluster / se_classical)

se_tbl |> arrange(desc(inflation_vs_classical)) |> head(12) |>
  kable(digits = 4,
        caption = "Classical vs HC3 vs country-clustered standard errors (top 12 by inflation)")

cat(sprintf("Median SE inflation from clustering: x%.2f | terms significant at 5%%: %d classical vs %d clustered\n",
            stats::median(se_tbl$inflation_vs_classical),
            sum(summary(m_full)$coefficients[-1, 4] < .05),
            sum(se_tbl$p_cluster < .05)))
```

## Normality of residuals

```{r diag-norm}
res <- stats::rstandard(m_full)
sw  <- stats::shapiro.test(sample(res, min(5000, length(res))))
tibble(test = "Shapiro-Wilk (sampled)", W = unname(sw$statistic), p.value = sw$p.value) |>
  kable(digits = 5, caption = "Normality check on standardised residuals")
```

## Influential observations


```{r diag-influence}
n_obs <- nrow(dat); p_par <- length(coef(m_full))

infl <- dat |>
  select(country, year, life_expectancy) |>
  mutate(
    cooksD   = stats::cooks.distance(m_full),
    leverage = stats::hatvalues(m_full),
    dffits   = stats::dffits(m_full),
    resid_std = stats::rstandard(m_full)
  )

thresholds <- tibble(
  measure   = c("Cook's D", "Leverage", "DFFITS"),
  threshold = c(4 / n_obs, 2 * p_par / n_obs, 2 * sqrt(p_par / n_obs)),
  n_exceeding = c(sum(infl$cooksD > 4 / n_obs),
                  sum(infl$leverage > 2 * p_par / n_obs),
                  sum(abs(infl$dffits) > 2 * sqrt(p_par / n_obs)))
)
kable(thresholds, digits = 4, caption = "Influence diagnostics and conventional cut-offs")

infl |> arrange(desc(cooksD)) |> head(15) |>
  kable(digits = 4, caption = "Fifteen most influential observations")
```

```{r diag-influence-country}
infl |>
  group_by(country) |>
  summarise(n_high_cook = sum(cooksD > 4 / n_obs), mean_cook = mean(cooksD),
            .groups = "drop") |>
  arrange(desc(n_high_cook)) |>
  head(12) |>
  kable(digits = 5, caption = "Countries contributing the most influential points")
```

```{r diag-dfbetas}
dfb <- as.data.frame(stats::dfbetas(m_full))
key_terms <- intersect(c("schooling", "adult_mortality", "log_hiv",
                         "income_composition"), names(dfb))

purrr::map_dfr(key_terms, function(tm) {
  i <- order(abs(dfb[[tm]]), decreasing = TRUE)[1:3]
  tibble(term = tm, country = dat$country[i], year = dat$year[i],
         dfbeta = dfb[[tm]][i])
}) |>
  kable(digits = 3,
        caption = "Observations that move key coefficients the most (top 3 per term)")
```

---

# Part 8 --- Multicollinearity

```{r vif, eval=PKG$car}
get_vif <- function(model) {
  v <- car::vif(model)
  if (is.matrix(v)) v <- v[, "GVIF^(1/(2*Df))"]^2   # comparable to a plain VIF
  tibble(term = names(v), VIF = as.numeric(v)) |> arrange(desc(VIF))
}

vif_full <- get_vif(m_full)
vif_full |>
  mutate(flag = dplyr::case_when(VIF > 10 ~ "severe", VIF > 5 ~ "moderate",
                                 TRUE ~ "")) |>
  kable(digits = 2, caption = "Variance inflation factors, full model")
```

```{r vif-reduce, eval=PKG$car}
drop_high_vif <- function(data, preds, fixed = c("status", "year_c"), thresh = 5) {
  keep <- preds
  repeat {
    f <- stats::reformulate(c(keep, fixed), response = "life_expectancy")
    v <- get_vif(lm(f, data = data)) |> filter(term %in% keep)
    if (nrow(v) == 0 || max(v$VIF) <= thresh) break
    worst <- v$term[which.max(v$VIF)]
    message("dropping ", worst, " (VIF = ", round(max(v$VIF), 1), ")")
    keep <- setdiff(keep, worst)
    if (length(keep) <= 2) break
  }
  keep
}

PRED_REDUCED <- drop_high_vif(dat, PRED_FULL, thresh = 5)
cat("Retained predictors:\n  ", paste(PRED_REDUCED, collapse = ", "), "\n")
cat("Removed for collinearity:\n  ",
    paste(setdiff(PRED_FULL, PRED_REDUCED), collapse = ", "), "\n")

form_red <- stats::reformulate(c(PRED_REDUCED, "status", "year_c"),
                               response = "life_expectancy")
m_red <- lm(form_red, data = dat)
get_vif(m_red) |> kable(digits = 2, caption = "VIF after backward removal")
```

```{r vif-fallback, eval=!PKG$car}
PRED_REDUCED <- setdiff(PRED_FULL, c("under_five_deaths", "thinness_5_9",
                                     "percentage_expenditure"))
form_red <- stats::reformulate(c(PRED_REDUCED, "status", "year_c"),
                               response = "life_expectancy")
m_red <- lm(form_red, data = dat)
cat("car not installed - using a fixed reduced predictor set.\n")
```

```{r vif-compare}
bind_rows(
  bind_cols(tibble(model = "Full"), eval_metrics(dat$life_expectancy, fitted(m_full)),
            tibble(k = length(coef(m_full)), AIC = AIC(m_full), BIC = BIC(m_full))),
  bind_cols(tibble(model = "Reduced (VIF-pruned)"),
            eval_metrics(dat$life_expectancy, fitted(m_red)),
            tibble(k = length(coef(m_red)), AIC = AIC(m_red), BIC = BIC(m_red)))
) |>
  kable(digits = 3, caption = "Dropping collinear predictors costs almost no fit")
```

```{r vif-cost}
r2_cost <- summary(m_full)$adj.r.squared - summary(m_red)$adj.r.squared
cat(sprintf("Adjusted R-squared lost by pruning: %.4f (%d predictors removed)\n",
            r2_cost, length(coef(m_full)) - length(coef(m_red))))
```
---

# Part 9 --- Interaction Analysis


```{r inter-fit}
# `a*b` is used rather than `a:b` so that main effects are always retained.
candidate_int <- c("schooling*status", "income_composition*status",
                   "log_hiv*status", "schooling*income_composition")

interaction_terms <- candidate_int[
  vapply(strsplit(candidate_int, "[*]"),
         function(v) all(trimws(v) %in% names(dat)), logical(1))
]

form_int <- stats::reformulate(c(PRED_REDUCED, "status", "year_c", interaction_terms),
                               response = "life_expectancy")
m_int <- lm(form_int, data = dat)

tidy_fit(m_int) |>
  filter(grepl(":", term)) |>
  mutate(sig = stars(p.value)) |>
  kable(digits = 4, caption = "Interaction terms only")
```

```{r inter-anova}
av <- stats::anova(m_red, m_int)
tibble(model    = c("Reduced (no interactions)", "With interactions"),
       res_df   = av$Res.Df,
       RSS      = av$RSS,
       F        = av$F,
       p.value  = av[["Pr(>F)"]]) |>
  kable(digits = 4,
        caption = "Classical nested F-test - shown for completeness, NOT used for inference")
```



```{r inter-mixed, eval=PKG$lme4}
lmer_fun0 <- if (PKG$lmerTest) lmerTest::lmer else lme4::lmer

int_terms_mm <- interaction_terms
form_mm_main <- paste(c(PRED_REDUCED, "status", "year_c"), collapse = " + ")

m_mm_main <- lmer_fun0(
  stats::as.formula(paste("life_expectancy ~", form_mm_main, "+ (1 | country)")),
  data = dat, REML = FALSE)

m_mm_int <- lmer_fun0(
  stats::as.formula(paste("life_expectancy ~", form_mm_main, "+",
                          paste(int_terms_mm, collapse = " + "), "+ (1 | country)")),
  data = dat, REML = FALSE)

as.data.frame(stats::anova(m_mm_main, m_mm_int)) |>
  tibble::rownames_to_column("model") |>
  kable(digits = 3,
        caption = "Likelihood-ratio test for the interactions, with a country random intercept")

fi <- as.data.frame(stats::coef(summary(m_mm_int)))
fi$term <- rownames(fi)
fi |> as_tibble() |> filter(grepl(":", term)) |> select(term, everything()) |>
  kable(digits = 4, caption = "Interaction terms estimated inside the mixed model")
```


```{r inter-plot, fig.height=4.4}
grid_schooling <- tidyr::expand_grid(
  schooling = seq(min(dat$schooling), max(dat$schooling), length.out = 60),
  status    = factor(c("Developing", "Developed"),
                     levels = levels(dat$status))
)
model_vars <- all.vars(stats::delete.response(stats::terms(m_int)))
need_vars  <- setdiff(model_vars, c("schooling", "status"))
others <- dat |>
  summarise(across(all_of(need_vars),
                   ~ if (is.numeric(.x)) stats::median(.x) else dplyr::first(.x)))
grid_schooling <- dplyr::bind_cols(grid_schooling,
                                   others[rep(1, nrow(grid_schooling)), , drop = FALSE])

grid_schooling$pred <- stats::predict(m_int, newdata = grid_schooling)

ggplot(grid_schooling, aes(schooling, pred, colour = status)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(values = pal_status) +
  labs(title = "Interaction: schooling x development status",
       subtitle = "Other predictors held at their medians; different slopes = interaction",
       x = "Years of schooling", y = "Predicted life expectancy", colour = NULL)
```

---

# Part 10 --- Model Selection


```{r selection, eval=PKG$MASS}
m_step <- MASS::stepAIC(m_int, direction = "both", trace = FALSE)
m_step_bic <- MASS::stepAIC(m_int, direction = "both", trace = FALSE,
                            k = log(nrow(dat)))

selection_tbl <- bind_rows(
  bind_cols(tibble(model = "Baseline"), eval_metrics(dat$life_expectancy, fitted(m_base)),
            tibble(k = length(coef(m_base)), AIC = AIC(m_base), BIC = BIC(m_base),
                   adj_R2 = summary(m_base)$adj.r.squared)),
  bind_cols(tibble(model = "Full"), eval_metrics(dat$life_expectancy, fitted(m_full)),
            tibble(k = length(coef(m_full)), AIC = AIC(m_full), BIC = BIC(m_full),
                   adj_R2 = summary(m_full)$adj.r.squared)),
  bind_cols(tibble(model = "Reduced (VIF)"), eval_metrics(dat$life_expectancy, fitted(m_red)),
            tibble(k = length(coef(m_red)), AIC = AIC(m_red), BIC = BIC(m_red),
                   adj_R2 = summary(m_red)$adj.r.squared)),
  bind_cols(tibble(model = "With interactions"), eval_metrics(dat$life_expectancy, fitted(m_int)),
            tibble(k = length(coef(m_int)), AIC = AIC(m_int), BIC = BIC(m_int),
                   adj_R2 = summary(m_int)$adj.r.squared)),
  bind_cols(tibble(model = "Stepwise (AIC)"), eval_metrics(dat$life_expectancy, fitted(m_step)),
            tibble(k = length(coef(m_step)), AIC = AIC(m_step), BIC = BIC(m_step),
                   adj_R2 = summary(m_step)$adj.r.squared)),
  bind_cols(tibble(model = "Stepwise (BIC)"), eval_metrics(dat$life_expectancy, fitted(m_step_bic)),
            tibble(k = length(coef(m_step_bic)), AIC = AIC(m_step_bic), BIC = BIC(m_step_bic),
                   adj_R2 = summary(m_step_bic)$adj.r.squared))
)
kable(selection_tbl, digits = 3,
      caption = "In-sample comparison. These numbers are optimistic by construction.")

cat("Formula chosen by AIC:\n"); print(stats::formula(m_step))
cat("\nFormula chosen by BIC:\n"); print(stats::formula(m_step_bic))
```

```{r selection-fallback, eval=!PKG$MASS}
m_step <- m_int; m_step_bic <- m_red
cat("MASS not installed - stepwise selection skipped.\n")
```


---

# Part 11 --- Regularization


## Folds respect the panel structure


```{r group-folds}
group_kfold <- function(groups, k = 5, seed = 2024) {
  set.seed(seed)
  ug   <- sample(unique(as.character(groups)))
  map  <- stats::setNames(rep_len(seq_len(k), length(ug)), ug)
  unname(map[as.character(groups)])
}

foldid     <- group_kfold(panel$country, k = CFG$n_folds, seed = CFG$seed)  # raw panel
foldid_dat <- group_kfold(dat$country,   k = CFG$n_folds, seed = CFG$seed)  # imputed panel

tibble(fold = foldid, country = panel$country) |>
  group_by(fold) |>
  summarise(countries = n_distinct(country), rows = n(), .groups = "drop") |>
  kable(caption = "Grouped folds: every country lives in exactly one fold")
```

```{r glmnet-data}
mm_form <- stats::reformulate(c(PRED_FULL, "status", "year_c"),
                              response = "life_expectancy")
X <- stats::model.matrix(mm_form, data = dat)[, -1, drop = FALSE]
colnames(X) <- make.names(colnames(X))
y <- dat$life_expectancy
cat("Design matrix:", nrow(X), "x", ncol(X), "\n")
```



```{r glmnet-fit, eval=PKG$glmnet, fig.height=4.6}
alphas <- c(0, 0.25, 0.5, 0.75, 1)

cv_list <- lapply(alphas, function(a) {
  glmnet::cv.glmnet(X, y, alpha = a, foldid = foldid_dat, standardize = TRUE)
})
names(cv_list) <- paste0("alpha_", alphas)

alpha_tbl <- purrr::imap_dfr(cv_list, function(cv, nm) {
  i <- which(cv$lambda == cv$lambda.min)
  tibble(alpha = as.numeric(sub("alpha_", "", nm)),
         lambda_min = cv$lambda.min, lambda_1se = cv$lambda.1se,
         cv_mse = cv$cvm[i], cv_rmse = sqrt(cv$cvm[i]),
         nonzero_at_1se = cv$nzero[which(cv$lambda == cv$lambda.1se)])
})
kable(alpha_tbl, digits = 4,
      caption = "Elastic-net mixing parameter selected by grouped cross-validation")

best_alpha <- alpha_tbl$alpha[which.min(alpha_tbl$cv_rmse)]
cat("Best alpha:", best_alpha, "\n")

op <- graphics::par(mfrow = c(1, 2), mar = c(4.2, 4.2, 3.4, 1))
plot(cv_list[["alpha_0"]]); graphics::title("Ridge", line = 2.6)
plot(cv_list[["alpha_1"]]); graphics::title("Lasso", line = 2.6)
graphics::par(op)
```

```{r glmnet-coefs, eval=PKG$glmnet, fig.height=6.5}
extract_coefs <- function(cv, label, s = "lambda.1se") {
  cf <- as.matrix(stats::coef(cv, s = s))
  tibble(term = rownames(cf), estimate = cf[, 1], model = label) |>
    filter(term != "(Intercept)")
}

coef_cmp <- bind_rows(
  extract_coefs(cv_list[["alpha_0"]], "Ridge"),
  extract_coefs(cv_list[[paste0("alpha_", best_alpha)]],
                paste0("Elastic net (a=", best_alpha, ")")),
  extract_coefs(cv_list[["alpha_1"]], "Lasso")
)

sd_x <- apply(X, 2, stats::sd)
coef_cmp |>
  mutate(std_estimate = estimate * sd_x[term]) |>   # per 1 SD of the predictor
  ggplot(aes(std_estimate, fct_reorder(term, abs(std_estimate)), fill = model)) +
  geom_col(position = position_dodge(width = .8), width = .75) +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("#1b6ca8", "#e07b39", "#5b8c5a")) +
  labs(title = "Standardised coefficients under three penalties",
       subtitle = "Effect of a one-standard-deviation change, at lambda.1se",
       x = "Change in life expectancy (years)", y = NULL, fill = NULL)

zeroed <- extract_coefs(cv_list[["alpha_1"]], "Lasso") |>
  filter(estimate == 0) |>
  pull(term)
cat("Predictors eliminated by the lasso at lambda.1se:\n  ",
    if (length(zeroed)) paste(zeroed, collapse = ", ") else "(none)", "\n")
```

---

# Part 12 --- Multilevel / Mixed-Effects Models


```{r mixed-prep, eval=PKG$lme4}
lmer_fun <- if (PKG$lmerTest) lmerTest::lmer else lme4::lmer

# z-scoring the predictors keeps the optimiser well behaved and makes fixed effects
# comparable (one SD of the predictor -> beta years of life expectancy)
Z_VARS <- c(PRED_REDUCED, "year_c")
dat_z <- dat |>
  mutate(across(all_of(Z_VARS), ~ as.numeric(scale(.x)), .names = "{.col}_z"))
Z_TERMS <- paste0(Z_VARS, "_z")

form_fixed <- paste(c(Z_TERMS, "status"), collapse = " + ")

# variable used for the random slope (falls back if VIF pruning removed schooling)
slope_var <- if ("schooling_z" %in% names(dat_z)) "schooling_z" else Z_TERMS[1]
```

## Random intercept

```{r mixed-ri, eval=PKG$lme4}
m_ri <- lmer_fun(
  stats::as.formula(paste("life_expectancy ~", form_fixed, "+ (1 | country)")),
  data = dat_z, REML = TRUE
)

fe <- as.data.frame(stats::coef(summary(m_ri)))
fe$term <- rownames(fe)
fe |>
  as_tibble() |>
  select(term, everything()) |>
  arrange(dplyr::desc(abs(Estimate))) |>
  kable(digits = 4, caption = "Random-intercept model: fixed effects (per 1 SD)")
```

```{r mixed-icc, eval=PKG$lme4}
vc <- as.data.frame(lme4::VarCorr(m_ri))
sig_u <- vc$vcov[vc$grp == "country"]
sig_e <- vc$vcov[vc$grp == "Residual"]

tibble(
  component = c("Between-country variance (sigma_u^2)",
                "Residual variance (sigma_e^2)",
                "Intraclass correlation (ICC)"),
  value = c(sig_u, sig_e, sig_u / (sig_u + sig_e))
) |>
  kable(digits = 4, caption = "Variance decomposition")
```


## Random slope

```{r mixed-rs, eval=PKG$lme4}
m_rs <- lmer_fun(
  stats::as.formula(paste0("life_expectancy ~ ", form_fixed,
                           " + (1 + ", slope_var, " | country)")),
  data = dat_z, REML = TRUE,
  control = lme4::lmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 2e5))
)

comp <- stats::anova(m_ri, m_rs)   # lme4 refits with ML for the LR test
as.data.frame(comp) |>
  tibble::rownames_to_column("model") |>
  kable(digits = 3, caption = "Likelihood-ratio test: random intercept vs random slope")
```

### Convergence and singularity

```{r mixed-singular, eval=PKG$lme4}
conv_report <- function(m, label) {
  msgs <- m@optinfo$conv$lme4$messages
  tibble(model = label,
         singular = lme4::isSingular(m, tol = 1e-4),
         n_theta = length(lme4::getME(m, "theta")),
         min_abs_theta = min(abs(lme4::getME(m, "theta"))),
         convergence_messages = if (is.null(msgs)) "none" else paste(msgs, collapse = " | "))
}

bind_rows(conv_report(m_ri, "Random intercept"),
          conv_report(m_rs, "Random intercept + slope")) |>
  kable(caption = "Convergence diagnostics (warnings are reported, not silenced)")

if (has("performance")) {
  cat("performance::check_singularity(m_rs):",
      performance::check_singularity(m_rs), "\n")
}

if (lme4::isSingular(m_rs, tol = 1e-4)) {
  cat("\nThe random-slope fit is SINGULAR. The random-intercept model is retained as\n",
      "the primary specification; the slope variance is not identified by these data.\n")
} else {
  cat("\nThe random-slope fit is not singular; the comparison above can be read normally.\n")
}
```

```{r mixed-ranef, eval=PKG$lme4, fig.height=4.4}
re <- lme4::ranef(m_ri, condVar = TRUE)$country
re_df <- tibble(country = rownames(re), intercept = re[, 1]) |>
  arrange(intercept) |>
  mutate(rank = dplyr::row_number())

bind_rows(head(re_df, 12), tail(re_df, 12)) |>
  mutate(country = fct_reorder(country, intercept),
         side = ifelse(intercept > 0, "above expectation", "below expectation")) |>
  ggplot(aes(intercept, country, fill = side)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("above expectation" = "#1b6ca8",
                               "below expectation" = "#c0392b")) +
  labs(title = "Country-specific residual baseline deviations",
       subtitle = "Estimated departure from the fixed-effects prediction, not a proven effect",
       x = "Random intercept (years)", y = NULL, fill = NULL)
```



```{r mixed-wb, eval=PKG$lme4}
wb_vars <- intersect(c("schooling", "income_composition", "log_gdp",
                       "adult_mortality"), PRED_REDUCED)
if (length(wb_vars) == 0)
  wb_vars <- intersect(c("schooling", "income_composition"), names(dat))

dat_wb <- dat |>
  group_by(country) |>
  mutate(across(all_of(wb_vars),
                list(btw = ~ mean(.x), wth = ~ .x - mean(.x)),
                .names = "{.col}_{.fn}")) |>
  ungroup()

wb_terms <- c(paste0(wb_vars, "_btw"), paste0(wb_vars, "_wth"))
m_wb <- lmer_fun(
  stats::as.formula(paste("life_expectancy ~", paste(wb_terms, collapse = " + "),
                          "+ status + year_c + (1 | country)")),
  data = dat_wb, REML = TRUE
)

fw <- as.data.frame(stats::coef(summary(m_wb)))
fw$term <- rownames(fw)
fw |>
  as_tibble() |>
  filter(grepl("_btw$|_wth$", term)) |>
  mutate(variable = sub("_(btw|wth)$", "", term),
         effect   = ifelse(grepl("_btw$", term), "between countries", "within country")) |>
  select(variable, effect, Estimate, `Std. Error`) |>
  arrange(variable, effect) |>
  kable(digits = 4, caption = "Between-country vs within-country associations")
```

---

# Part 13 --- Time Effects


```{r time-models, eval=PKG$lme4}
ctrl <- lme4::lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))

m_t_lin  <- lmer_fun(life_expectancy ~ year_c + (1 | country),
                     data = dat, REML = FALSE)
m_t_quad <- lmer_fun(life_expectancy ~ year_c + I(year_c^2) + (1 | country),
                     data = dat, REML = FALSE)
m_t_rs   <- lmer_fun(life_expectancy ~ year_c + (1 + year_c | country),
                     data = dat, REML = FALSE, control = ctrl)
m_t_full <- lmer_fun(life_expectancy ~ year_c + I(year_c^2) + (1 + year_c | country),
                     data = dat, REML = FALSE, control = ctrl)
```

### Pairwise nested tests only

```{r time-nested, eval=PKG$lme4}
lr_pair <- function(small, large, label) {
  a <- stats::anova(small, large)
  tibble(comparison = label, df_diff = a$Df[2], Chisq = a$Chisq[2],
         p.value = a[["Pr(>Chisq)"]][2])
}

bind_rows(
  lr_pair(m_t_lin,  m_t_quad, "linear -> quadratic time (fixed part)"),
  lr_pair(m_t_lin,  m_t_rs,   "common slope -> country-specific slope (random part)"),
  lr_pair(m_t_quad, m_t_full, "quadratic: common slope -> country-specific slope"),
  lr_pair(m_t_rs,   m_t_full, "country-specific slope: linear -> quadratic time")
) |>
  kable(digits = 4, caption = "Each row is a genuinely nested comparison")
```

### Non-nested comparison by information criterion

```{r time-ic, eval=PKG$lme4}
ic_row <- function(m, label) tibble(model = label, df = attr(stats::logLik(m), "df"),
                                    logLik = as.numeric(stats::logLik(m)),
                                    AIC = stats::AIC(m), BIC = stats::BIC(m),
                                    singular = lme4::isSingular(m, tol = 1e-4))
bind_rows(
  ic_row(m_t_lin,  "year + (1|country)"),
  ic_row(m_t_quad, "year + year^2 + (1|country)"),
  ic_row(m_t_rs,   "year + (1 + year|country)"),
  ic_row(m_t_full, "year + year^2 + (1 + year|country)")
) |>
  arrange(AIC) |>
  kable(digits = 2,
        caption = "AIC/BIC comparison, which does not require nesting (singular fits flagged)")
```

```{r time-slopes, eval=PKG$lme4, fig.height=4.2}
sl <- lme4::ranef(m_t_rs)$country
slopes <- tibble(country = rownames(sl),
                 slope = sl[["year_c"]] + lme4::fixef(m_t_rs)[["year_c"]])

ggplot(slopes, aes(slope)) +
  geom_histogram(bins = 40, fill = "#1b6ca8", colour = "white") +
  geom_vline(xintercept = lme4::fixef(m_t_rs)[["year_c"]], linetype = 2,
             colour = "#c0392b", linewidth = .9) +
  labs(title = "Countries improve at very different rates",
       subtitle = "Dashed line: the average (fixed-effect) annual improvement",
       x = "Estimated annual change in life expectancy (years/year)", y = "Countries")

bind_rows(
  slopes |> arrange(desc(slope)) |> head(6) |> mutate(group = "fastest improvement"),
  slopes |> arrange(slope) |> head(6) |> mutate(group = "slowest / declining")
) |>
  kable(digits = 3, caption = "Extremes of the country-specific time trend")
```

```{r time-quad, eval=PKG$lme4}
tibble(term = names(lme4::fixef(m_t_quad)),
       estimate = unname(lme4::fixef(m_t_quad))) |>
  kable(digits = 4, caption = "Quadratic time trend (global improvement, decelerating)")
```

---

# Part 14 --- Machine Learning Models

```{r ml-grids}
RF_GRID <- tidyr::expand_grid(
  mtry_frac = c(1/3, 1/2),
  min_node  = c(5, 20)
)

XGB_GRID <- tidyr::expand_grid(
  max_depth        = c(3, 5, 7),
  eta              = 0.05,
  min_child_weight = 3
)

XGB_FIXED <- list(objective = "reg:squarederror", subsample = 0.8,
                  colsample_bytree = 0.8)

# defaults used when CFG$tune_ml is FALSE
RF_DEFAULT  <- list(mtry_frac = 1/3, min_node = 5)
XGB_DEFAULT <- list(params = c(XGB_FIXED, list(eta = 0.05, max_depth = 5,
                                               min_child_weight = 3)),
                    nrounds = 600)

kable(RF_GRID, caption = "Random-forest tuning grid (inner CV)")
kable(XGB_GRID, caption = "XGBoost tuning grid (inner CV; nrounds by early stopping)")
```

```{r ml-tuners}
tune_rf <- function(tr, Xtr, ytr, inner_k = 3) {
  if (!CFG$tune_ml) return(RF_DEFAULT)
  fid <- group_kfold(tr$country, k = inner_k, seed = CFG$seed + 7)
  scores <- vapply(seq_len(nrow(RF_GRID)), function(g) {
    mtry <- max(1, floor(ncol(Xtr) * RF_GRID$mtry_frac[g]))
    mean(vapply(sort(unique(fid)), function(f) {
      i1 <- fid != f; i2 <- fid == f
      fit <- ranger::ranger(x = as.data.frame(Xtr[i1, , drop = FALSE]), y = ytr[i1],
                            num.trees = 300, mtry = mtry,
                            min.node.size = RF_GRID$min_node[g],
                            seed = CFG$seed, num.threads = 2)
      rmse(ytr[i2], stats::predict(fit, data = as.data.frame(Xtr[i2, , drop = FALSE]))$predictions)
    }, numeric(1)))
  }, numeric(1))
  as.list(RF_GRID[which.min(scores), ])
}

tune_xgb <- function(tr, Xtr, ytr, inner_k = 3) {
  if (!CFG$tune_ml) return(XGB_DEFAULT)
  fid   <- group_kfold(tr$country, k = inner_k, seed = CFG$seed + 7)
  folds <- split(seq_along(fid), fid)
  dtrain <- xgboost::xgb.DMatrix(data = as.matrix(Xtr), label = ytr)

  res <- lapply(seq_len(nrow(XGB_GRID)), function(g) {
    prm <- c(XGB_FIXED, list(eta = XGB_GRID$eta[g], max_depth = XGB_GRID$max_depth[g],
                             min_child_weight = XGB_GRID$min_child_weight[g]))
    cv <- xgboost::xgb.cv(params = prm, data = dtrain, nrounds = 1200, folds = folds,
                          early_stopping_rounds = 30, verbose = 0, nthread = 2)
    log_col <- grep("test.*rmse.*mean", names(cv$evaluation_log), value = TRUE)[1]
    list(params = prm, nrounds = cv$best_iteration,
         score = cv$evaluation_log[[log_col]][cv$best_iteration])
  })
  res[[which.min(vapply(res, function(z) z$score, numeric(1)))]]
}
```

---

# Part 15 --- Validation Strategy

## The pipeline, and why the order matters


```{r preprocess-fold}
finalize <- function(tr, te) {
  # backstop: a variable that was entirely missing in the training fold has no median.
  med <- vapply(tr[IMPUTE_VARS], function(z) stats::median(z, na.rm = TRUE), numeric(1))
  med[!is.finite(med)] <- 0
  for (v in IMPUTE_VARS) {
    tr[[v]][is.na(tr[[v]])] <- med[[v]]
    te[[v]][is.na(te[[v]])] <- med[[v]]
  }
  relog <- function(d) d |>
    mutate(log_gdp = log1p_safe(gdp), log_population = log1p_safe(population),
           log_measles = log1p_safe(measles), log_hiv = log1p_safe(hiv_aids))
  list(tr = relog(tr), te = relog(te))
}

build_matrix <- function(tr, te) {
  Xtr <- stats::model.matrix(mm_form, data = tr)[, -1, drop = FALSE]
  Xte <- stats::model.matrix(mm_form, data = te)[, -1, drop = FALSE]
  add <- setdiff(colnames(Xtr), colnames(Xte))
  if (length(add)) {
    pad <- matrix(0, nrow(Xte), length(add), dimnames = list(NULL, add))
    Xte <- cbind(Xte, pad)
  }
  Xte <- Xte[, colnames(Xtr), drop = FALSE]
  colnames(Xtr) <- make.names(colnames(Xtr))
  colnames(Xte) <- colnames(Xtr)
  list(Xtr = Xtr, Xte = Xte)
}

# THE fold preprocessor: everything it knows, it learned from train_raw.
preprocess_fold <- function(train_raw, test_raw, test_mode = "interior") {
  prm <- impute_fit(train_raw, IMPUTE_VARS)
  tr  <- impute_apply(train_raw, prm, mode = "interior")
  te  <- impute_apply(test_raw, prm,
                      history = if (test_mode == "causal") train_raw else NULL,
                      mode    = test_mode)
  f <- finalize(tr, te)
  M <- build_matrix(f$tr, f$te)
  list(tr = f$tr, te = f$te, Xtr = M$Xtr, Xte = M$Xte,
       ytr = f$tr$life_expectancy, yte = f$te$life_expectancy)
}
```


```{r cv-engine}
run_cv <- function(folds, specs, data = panel, test_mode = "interior") {
  out <- list()
  for (f in sort(unique(folds))) {
    itr <- which(folds != f); ite <- which(folds == f)
    P <- preprocess_fold(data[itr, ], data[ite, ],
                         test_mode = if (CFG$test_interpolate) test_mode else "causal")
    for (nm in names(specs)) {
      pred <- tryCatch(specs[[nm]](P),
        error = function(e) { warning(nm, " failed on fold ", f, ": ",
                                      conditionMessage(e)); rep(NA_real_, length(ite)) })
      out[[length(out) + 1]] <- tibble(model = nm, fold = f,
                                       obs = P$yte, pred = as.numeric(pred))
    }
  }
  bind_rows(out)
}

summarise_cv <- function(preds) {
  per_fold <- preds |>
    group_by(model, fold) |>
    summarise(RMSE = rmse(obs, pred), MAE = mae(obs, pred), R2 = r2(obs, pred),
              .groups = "drop")
  fold_avg <- per_fold |>
    group_by(model) |>
    summarise(RMSE_foldmean = mean(RMSE), RMSE_sd = stats::sd(RMSE),
              MAE_foldmean = mean(MAE), R2_foldmean = mean(R2), .groups = "drop")
  # pooled out-of-fold metrics: every observation predicted exactly once
  pooled <- preds |>
    group_by(model) |>
    summarise(RMSE_oof = rmse(obs, pred), MAE_oof = mae(obs, pred),
              R2_oof = r2(obs, pred), .groups = "drop")
  left_join(pooled, fold_avg, by = "model") |> arrange(RMSE_oof)
}
```

```{r cv-specs}
scale_train <- function(Xtr, Xte) {
  mu <- colMeans(Xtr); sdv <- apply(Xtr, 2, stats::sd); sdv[sdv == 0] <- 1
  list(tr = sweep(sweep(Xtr, 2, mu, "-"), 2, sdv, "/"),
       te = sweep(sweep(Xte, 2, mu, "-"), 2, sdv, "/"))
}

make_specs <- function(use_random_effects = FALSE) {
  s <- list(
    "Simple regression" = function(P)
      stats::predict(lm(life_expectancy ~ schooling, data = P$tr), newdata = P$te),

    "Multiple regression" = function(P)
      stats::predict(lm(form_red, data = P$tr), newdata = P$te),

    "MLR + interactions" = function(P)
      stats::predict(lm(form_int, data = P$tr), newdata = P$te)
  )

  if (PKG$lme4) {
    s[["Mixed effects (RI)"]] <- function(P) {
      sc  <- scale_train(P$Xtr, P$Xte)
      dtr <- as.data.frame(sc$tr); dtr$life_expectancy <- P$ytr; dtr$country <- P$tr$country
      dte <- as.data.frame(sc$te); dte$country <- P$te$country
      fml <- stats::as.formula(paste("life_expectancy ~",
                                     paste(colnames(sc$tr), collapse = " + "),
                                     "+ (1 | country)"))
      m <- suppressWarnings(suppressMessages(
        lme4::lmer(fml, data = dtr, REML = TRUE,
                   control = lme4::lmerControl(optimizer = "bobyqa",
                                               optCtrl = list(maxfun = 2e5)))))
      stats::predict(m, newdata = dte, allow.new.levels = TRUE,
                     re.form = if (use_random_effects) NULL else NA)
    }
  }

  if (PKG$glmnet) {
    penalised <- function(alpha) function(P) {
      inner <- group_kfold(P$tr$country, k = 5, seed = CFG$seed + 1)
      cv <- glmnet::cv.glmnet(P$Xtr, P$ytr, alpha = alpha, foldid = inner,
                              standardize = TRUE)
      as.numeric(stats::predict(cv, newx = P$Xte, s = "lambda.min"))
    }
    s[["Ridge"]]       <- penalised(0)
    s[["Lasso"]]       <- penalised(1)
    s[["Elastic net"]] <- penalised(0.5)
  }

  if (PKG$ranger) {
    s[["Random forest"]] <- function(P) {
      hp  <- tune_rf(P$tr, P$Xtr, P$ytr)
      fit <- ranger::ranger(x = as.data.frame(P$Xtr), y = P$ytr,
                            num.trees = CFG$rf_trees,
                            mtry = max(1, floor(ncol(P$Xtr) * hp$mtry_frac)),
                            min.node.size = hp$min_node,
                            seed = CFG$seed, num.threads = 2)
      stats::predict(fit, data = as.data.frame(P$Xte))$predictions
    }
  }

  if (PKG$xgboost && CFG$run_xgboost) {
    s[["XGBoost"]] <- function(P) {
      hp  <- tune_xgb(P$tr, P$Xtr, P$ytr)
      bst <- xgboost::xgb.train(params = hp$params,
                                data = xgboost::xgb.DMatrix(as.matrix(P$Xtr), label = P$ytr),
                                nrounds = hp$nrounds, verbose = 0, nthread = 2)
      stats::predict(bst, as.matrix(P$Xte))
    }
  }
  s
}

specs_grouped <- make_specs(use_random_effects = FALSE)
names(specs_grouped)
```

## Evaluation A --- grouped cross-validation over countries

```{r cv-grouped, cache=TRUE}
cv_preds   <- run_cv(foldid, specs_grouped, data = panel, test_mode = "interior")
cv_summary <- summarise_cv(cv_preds)
kable(cv_summary, digits = 3,
      caption = "Grouped CV: every test country is unseen, and its preprocessing was never fitted on it")
```



```{r cv-grouped-plot, fig.height=4.6}
cv_preds |>
  group_by(model, fold) |>
  summarise(RMSE = rmse(obs, pred), .groups = "drop") |>
  mutate(model = fct_reorder(model, RMSE)) |>
  ggplot(aes(RMSE, model)) +
  geom_boxplot(fill = "#1b6ca8", alpha = .3, outlier.shape = NA) +
  geom_jitter(height = .12, alpha = .8, colour = "#1b6ca8") +
  labs(title = "Fold-to-fold variability, grouped CV",
       subtitle = "One point per fold; spread matters as much as the mean",
       x = "RMSE (years)", y = NULL)
```

## Evaluation B --- temporal split

```{r cv-temporal, cache=TRUE}
train_idx <- which(panel$year <= CFG$temporal_cutoff)
test_idx  <- which(panel$year >  CFG$temporal_cutoff)

cat("Train:", length(train_idx), "rows (<=", CFG$temporal_cutoff, ") | Test:",
    length(test_idx), "rows (>", CFG$temporal_cutoff, ")\n")

# causal mode: the test transform never looks forward in time
P_temp <- preprocess_fold(panel[train_idx, ], panel[test_idx, ], test_mode = "causal")

# here the test countries HAVE been seen, so the random effects are usable
specs_temporal <- make_specs(use_random_effects = TRUE)

temporal_preds <- purrr::imap_dfr(specs_temporal, function(fn, nm) {
  pred <- tryCatch(fn(P_temp),
    error = function(e) { warning(nm, ": ", conditionMessage(e))
                          rep(NA_real_, length(test_idx)) })
  tibble(model = nm, obs = P_temp$yte, pred = as.numeric(pred))
})

temporal_summary <- temporal_preds |>
  group_by(model) |>
  summarise(RMSE = rmse(obs, pred), MAE = mae(obs, pred), R2 = r2(obs, pred),
            .groups = "drop") |>
  arrange(RMSE)
kable(temporal_summary, digits = 3,
      caption = "Temporal validation: train on 2000-2012, predict 2013-2015 (causal preprocessing)")
```


```{r cv-naive, cache=TRUE}
naive_folds <- sample(rep_len(seq_len(CFG$n_folds), nrow(panel)))
naive_specs <- specs_grouped[intersect(c("Multiple regression", "Random forest"),
                                       names(specs_grouped))]
naive_summary <- summarise_cv(run_cv(naive_folds, naive_specs, data = panel))

naive_cmp <- bind_rows(
  naive_summary |> mutate(scheme = "Random row split (WRONG)"),
  cv_summary |> filter(model %in% naive_summary$model) |>
    mutate(scheme = "Grouped by country (correct)")
) |>
  select(scheme, model, RMSE_oof, MAE_oof, R2_oof) |>
  arrange(model, scheme)

kable(naive_cmp, digits = 3,
      caption = "The optimism produced by ignoring the panel structure")

naive_gap <- naive_cmp |>
  group_by(model) |>
  summarise(gap = RMSE_oof[scheme == "Grouped by country (correct)"] -
                  RMSE_oof[scheme == "Random row split (WRONG)"], .groups = "drop")
```

---

# Part 16 --- Model Comparison

```{r comparison}
interpretability <- tibble(
  model = c("Simple regression", "Multiple regression", "MLR + interactions",
            "Mixed effects (RI)", "Ridge", "Lasso", "Elastic net",
            "Random forest", "XGBoost"),
  interpretability = c("Very high", "High", "High", "High", "Medium",
                       "Medium-high", "Medium", "Medium", "Medium"),
  handles_panel = c("No", "No", "No", "Yes", "No", "No", "No", "No", "No"),
  valid_inference = c("Only with clustered SEs", "Only with clustered SEs",
                      "Only with clustered SEs", "Yes", "Limited", "Limited",
                      "Limited", "No", "No")
)

final_table <- cv_summary |>
  select(model, cv_RMSE = RMSE_oof, cv_MAE = MAE_oof, cv_R2 = R2_oof,
         cv_RMSE_sd = RMSE_sd) |>
  left_join(temporal_summary |>
              select(model, temporal_RMSE = RMSE, temporal_MAE = MAE,
                     temporal_R2 = R2), by = "model") |>
  left_join(interpretability, by = "model") |>
  arrange(cv_RMSE)

kable(final_table, digits = 3,
      caption = "Central comparison table: pooled out-of-fold metrics, temporal metrics, interpretability")
```

```{r comparison-dynamic}
best_grouped  <- final_table |> slice_min(cv_RMSE, n = 1, with_ties = FALSE) |> pull(model)
best_temporal <- temporal_summary |> slice_min(RMSE, n = 1, with_ties = FALSE) |> pull(model)
best_linear   <- final_table |>
  filter(model %in% c("Multiple regression", "MLR + interactions", "Ridge",
                      "Lasso", "Elastic net")) |>
  slice_min(cv_RMSE, n = 1, with_ties = FALSE) |> pull(model)
tree_models   <- c("Random forest", "XGBoost")
trees_lead    <- best_grouped %in% tree_models
same_winner   <- identical(best_grouped, best_temporal)
```


```{r comparison-plot, fig.height=4.6}
final_table |>
  select(model, `Grouped CV` = cv_RMSE, `Temporal` = temporal_RMSE) |>
  pivot_longer(-model, names_to = "scheme", values_to = "RMSE") |>
  filter(!is.na(RMSE)) |>
  mutate(model = fct_reorder(model, RMSE)) |>
  ggplot(aes(RMSE, model, fill = scheme)) +
  geom_col(position = position_dodge(width = .8), width = .72) +
  scale_fill_manual(values = c("Grouped CV" = "#1b6ca8", "Temporal" = "#e07b39")) +
  labs(title = "Out-of-sample error under both validation schemes",
       subtitle = "Pooled out-of-fold RMSE (grouped) and hold-out RMSE (temporal)",
       x = "RMSE (years)", y = NULL, fill = NULL)
```

```{r comparison-residuals, fig.height=4.6}
cv_preds |>
  filter(model %in% unique(c("Multiple regression", best_grouped))) |>
  ggplot(aes(pred, obs)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey40", linetype = 2) +
  geom_point(alpha = .2, size = .8, colour = "#1b6ca8") +
  facet_wrap(~ model) +
  labs(title = "Predicted vs observed on held-out countries",
       x = "Predicted life expectancy", y = "Observed life expectancy")
```

---

# Part 17 --- Interpretability

```{r interp-holdout}
h_tr <- which(foldid != 1); h_te <- which(foldid == 1)
P1   <- preprocess_fold(panel[h_tr, ], panel[h_te, ], test_mode = "interior")

cat("Holdout: fold 1 -", length(h_te), "rows,",
    n_distinct(panel$country[h_te]), "countries\n")
```

## Permutation importance

A predictor's importance is the *increase in holdout RMSE* when its values are shuffled,
which breaks its relationship with the outcome while preserving its marginal distribution.

```{r interp-perm, eval=PKG$ranger, fig.height=6}
hp_rf  <- tune_rf(P1$tr, P1$Xtr, P1$ytr)
rf_fit <- ranger::ranger(x = as.data.frame(P1$Xtr), y = P1$ytr,
                         num.trees = CFG$rf_trees,
                         mtry = max(1, floor(ncol(P1$Xtr) * hp_rf$mtry_frac)),
                         min.node.size = hp_rf$min_node,
                         importance = "permutation", seed = CFG$seed, num.threads = 2)

rf_pred_fun <- function(newX)
  stats::predict(rf_fit, data = as.data.frame(newX))$predictions

cat("Holdout RMSE:", round(rmse(P1$yte, rf_pred_fun(P1$Xte)), 3), "years\n")

perm_importance <- function(pred_fun, Xte, yte, n_rep = 5, seed = CFG$seed) {
  set.seed(seed)
  base <- rmse(yte, pred_fun(Xte))
  vals <- vapply(colnames(Xte), function(v) {
    mean(replicate(n_rep, {
      Xp <- Xte; Xp[, v] <- sample(Xp[, v]); rmse(yte, pred_fun(Xp))
    }))
  }, numeric(1))
  tibble(variable = colnames(Xte), baseline_rmse = base,
         permuted_rmse = as.numeric(vals), increase = as.numeric(vals) - base) |>
    arrange(desc(increase))
}

perm_imp <- perm_importance(rf_pred_fun, P1$Xte, P1$yte)
kable(head(perm_imp, 12), digits = 4,
      caption = "Permutation importance on held-out countries (random forest)")

perm_imp |>
  mutate(variable = fct_reorder(variable, increase)) |>
  ggplot(aes(increase, variable)) +
  geom_col(fill = "#1b6ca8", alpha = .9) +
  labs(title = "Which predictors actually carry the prediction?",
       subtitle = "Increase in holdout RMSE when the predictor is shuffled",
       x = "RMSE increase (years)", y = NULL)
```

## Partial dependence

```{r interp-pdp, eval=PKG$ranger, fig.height=5}
partial_dep <- function(pred_fun, Xref, var, grid_n = 25, n_sample = 400) {
  set.seed(CFG$seed)
  idx <- sample(nrow(Xref), min(n_sample, nrow(Xref)))
  Xs  <- Xref[idx, , drop = FALSE]
  gr  <- stats::quantile(Xref[, var], probs = seq(.02, .98, length.out = grid_n),
                         names = FALSE)
  purrr::map_dfr(gr, function(v) {
    Xt <- Xs; Xt[, var] <- v
    tibble(variable = var, x = v, yhat = mean(pred_fun(Xt)))
  })
}

top_vars <- head(perm_imp$variable, 4)
pdp_df   <- purrr::map_dfr(top_vars, ~ partial_dep(rf_pred_fun, P1$Xtr, .x))

ggplot(pdp_df, aes(x, yhat)) +
  geom_line(colour = "#1b6ca8", linewidth = 1.1) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Partial dependence for the four strongest predictors",
       subtitle = "Average model prediction as one predictor is varied, others held as observed",
       x = NULL, y = "Predicted life expectancy (years)")
```


## SHAP values (exact, for the boosted trees)

```{r interp-shap, eval=PKG$xgboost && CFG$run_xgboost, fig.height=5.5}
hp_x <- tune_xgb(P1$tr, P1$Xtr, P1$ytr)
bst  <- xgboost::xgb.train(params = hp_x$params,
                           data = xgboost::xgb.DMatrix(as.matrix(P1$Xtr), label = P1$ytr),
                           nrounds = hp_x$nrounds, verbose = 0, nthread = 2)

shap <- stats::predict(bst, as.matrix(P1$Xte), predcontrib = TRUE)
colnames(shap) <- c(colnames(P1$Xtr), "BIAS")
shap_feat <- shap[, colnames(P1$Xtr), drop = FALSE]

shap_imp <- tibble(variable = colnames(shap_feat),
                   mean_abs_shap = colMeans(abs(shap_feat))) |>
  arrange(desc(mean_abs_shap))

kable(head(shap_imp, 12), digits = 4,
      caption = "Mean |SHAP| on held-out countries (XGBoost)")

shap_imp |>
  mutate(variable = fct_reorder(variable, mean_abs_shap)) |>
  ggplot(aes(mean_abs_shap, variable)) +
  geom_col(fill = "#5b8c5a", alpha = .9) +
  labs(title = "Average contribution of each predictor to individual predictions",
       x = "Mean |SHAP| (years)", y = NULL)
```

```{r interp-shap-dep, eval=PKG$xgboost && CFG$run_xgboost, fig.height=4.2}
v1 <- shap_imp$variable[1]
tibble(value = P1$Xte[, v1], shap = shap_feat[, v1],
       status = P1$te$status) |>
  ggplot(aes(value, shap, colour = status)) +
  geom_hline(yintercept = 0, colour = "grey45") +
  geom_point(alpha = .45, size = 1) +
  geom_smooth(se = FALSE, colour = "black", linewidth = .7) +
  scale_colour_manual(values = pal_status) +
  labs(title = paste("SHAP dependence:", v1),
       subtitle = "Contribution of this predictor to each individual prediction",
       x = v1, y = "SHAP value (years)", colour = NULL)
```


```{r interp-agree, eval=PKG$ranger}
rank_tbl <- perm_imp |> transmute(variable, rf_rank = dplyr::row_number())

if (exists("shap_imp")) {
  rank_tbl <- rank_tbl |>
    left_join(shap_imp |> transmute(variable, xgb_rank = dplyr::row_number()),
              by = "variable")
}

if (PKG$glmnet && exists("cv_list") && exists("sd_x")) {
  lin_rank <- extract_coefs(cv_list[["alpha_1"]], "Lasso") |>
    mutate(term = make.names(term),
           std_abs = abs(estimate * sd_x[match(term, names(sd_x))])) |>
    filter(!is.na(std_abs)) |>
    arrange(desc(std_abs)) |>
    transmute(variable = term, lasso_rank = dplyr::row_number())
  rank_tbl <- rank_tbl |> left_join(lin_rank, by = "variable")
}

head(rank_tbl, 12) |>
  kable(caption = "Importance rank under the definitions available in this run")
```

---

# Part 18 --- Statistical vs Predictive Conclusions

```{r conclusions-table}
tibble(
  question = c("Statistical / inferential", "Predictive"),
  asks = c("Which variables are associated with life expectancy, and by how much?",
           "Which model best predicts observations it has never seen?"),
  best_tool = c("Mixed-effects model (random intercept per country)",
                paste("Grouped CV:", best_grouped, "| Temporal:", best_temporal)),
  why = c("Correct standard errors under repeated measures; separates within- from between-country effects",
          "Lowest out-of-fold error; captures non-linearity without specifying it")
) |>
  kable(caption = "Two questions, two winners")
```

---

# Part 19 --- Secondary Classification Analysis

```{r clf-setup, eval=PKG$MASS}
FIXED_CUTS <- c(65, 75)   # substantive alternative; set USE_FIXED_CUTS <- TRUE to apply
USE_FIXED_CUTS <- FALSE

to_class <- function(y, cuts)
  cut(y, breaks = c(-Inf, cuts, Inf), labels = c("Low", "Medium", "High"))

CLF_VARS <- PRED_REDUCED

macro_f1 <- function(obs, pred) {
  lv <- levels(obs)
  mean(vapply(lv, function(l) {
    tp <- sum(pred == l & obs == l); fp <- sum(pred == l & obs != l)
    fn <- sum(pred != l & obs == l)
    if (tp == 0) return(0)
    p <- tp / (tp + fp); r <- tp / (tp + fn); 2 * p * r / (p + r)
  }, numeric(1)))
}

clf_specs <- list()
if (PKG$nnet) {
  clf_specs[["Multinomial logistic"]] <- function(tr_x, tr_y, te_x) {
    d <- as.data.frame(tr_x); d$.y <- tr_y
    m <- nnet::multinom(.y ~ ., data = d, trace = FALSE, maxit = 400)
    stats::predict(m, newdata = as.data.frame(te_x))
  }
}
clf_specs[["LDA"]] <- function(tr_x, tr_y, te_x) {
  stats::predict(MASS::lda(x = tr_x, grouping = tr_y), te_x)$class
}
clf_specs[["QDA"]] <- function(tr_x, tr_y, te_x) {
  stats::predict(MASS::qda(x = tr_x, grouping = tr_y), te_x)$class
}
if (PKG$class) {
  clf_specs[["KNN (k=15)"]] <- function(tr_x, tr_y, te_x)
    class::knn(train = tr_x, test = te_x, cl = tr_y, k = 15)
}
```

```{r clf-cv, eval=PKG$MASS, cache=TRUE}
clf_results <- purrr::map_dfr(sort(unique(foldid)), function(f) {
  itr <- which(foldid != f); ite <- which(foldid == f)
  P <- preprocess_fold(panel[itr, ], panel[ite, ], test_mode = "interior")

  # cut-points learned on the TRAINING outcomes only
  cuts <- if (USE_FIXED_CUTS) FIXED_CUTS else
    unname(stats::quantile(P$ytr, probs = c(1/3, 2/3)))
  ytr_c <- to_class(P$ytr, cuts)
  yte_c <- to_class(P$yte, cuts)

  Xtr <- as.matrix(P$tr[, CLF_VARS]); Xte <- as.matrix(P$te[, CLF_VARS])
  mu <- colMeans(Xtr); sdv <- apply(Xtr, 2, stats::sd); sdv[sdv == 0] <- 1
  ztr <- sweep(sweep(Xtr, 2, mu, "-"), 2, sdv, "/")
  zte <- sweep(sweep(Xte, 2, mu, "-"), 2, sdv, "/")

  purrr::imap_dfr(clf_specs, function(fn, nm) {
    pr <- tryCatch(fn(ztr, ytr_c, zte),
                   error = function(e) { warning(nm, " fold ", f, ": ",
                                                 conditionMessage(e)); NULL })
    if (is.null(pr)) return(tibble())
    pr <- factor(pr, levels = levels(ytr_c))
    tibble(model = nm, fold = f,
           lower_cut = cuts[1], upper_cut = cuts[2],
           accuracy = mean(pr == yte_c),
           macro_f1 = macro_f1(yte_c, pr),
           adjacent_error = mean(abs(as.integer(pr) - as.integer(yte_c)) == 1),
           far_error = mean(abs(as.integer(pr) - as.integer(yte_c)) == 2))
  })
})

clf_results |>
  group_by(model) |>
  summarise(accuracy = mean(accuracy), accuracy_sd = stats::sd(accuracy),
            macro_f1 = mean(macro_f1),
            adjacent_error = mean(adjacent_error), far_error = mean(far_error),
            .groups = "drop") |>
  arrange(desc(accuracy)) |>
  kable(digits = 3,
        caption = "Grouped CV with fold-wise cut-points (no label leakage)")

cat(sprintf("Fold-wise lower cut-points: %s\nFold-wise upper cut-points: %s\n",
            paste(round(unique(clf_results$lower_cut), 1), collapse = ", "),
            paste(round(unique(clf_results$upper_cut), 1), collapse = ", ")))
```

```{r clf-verdict}
err_split <- if (exists("clf_results") && nrow(clf_results) > 0) {
  clf_results |> summarise(adjacent = mean(adjacent_error), far = mean(far_error))
} else tibble(adjacent = NA_real_, far = NA_real_)
```
---

# Part 20 --- Robustness and Sensitivity Analysis

## Sensitivity 1 --- imputation strategy

```{r sens-imputation}
m_red_cc  <- lm(form_red, data = dat_cc)
m_red_imp <- lm(form_red, data = dat_imp)

sens_imp <- bind_rows(
  tidy_fit(m_red_cc,  conf = FALSE) |> mutate(dataset = "Complete cases"),
  tidy_fit(m_red_imp, conf = FALSE) |> mutate(dataset = "Panel-aware imputation")
)

sens_wide <- sens_imp |>
  select(term, dataset, estimate) |>
  pivot_wider(names_from = dataset, values_from = estimate) |>
  mutate(difference = `Panel-aware imputation` - `Complete cases`,
         rel_change_pct = 100 * difference / abs(`Complete cases`),
         sign_flip = sign(`Panel-aware imputation`) != sign(`Complete cases`))

sens_wide |> arrange(desc(abs(rel_change_pct))) |>
  kable(digits = 4, caption = "Coefficient stability across missing-data strategies")

cat(sprintf("Coefficients that change sign between the two strategies: %d of %d\n",
            sum(sens_wide$sign_flip, na.rm = TRUE), nrow(sens_wide)))
```

```{r sens-imputation-plot, fig.height=5.5}
sens_imp |>
  filter(term != "(Intercept)") |>
  ggplot(aes(estimate, fct_reorder(term, estimate), colour = dataset)) +
  geom_vline(xintercept = 0, colour = "grey45", linetype = 2) +
  geom_point(size = 2.4, position = position_dodge(width = .5)) +
  scale_colour_manual(values = c("Complete cases" = "#c0392b",
                                 "Panel-aware imputation" = "#1b6ca8")) +
  labs(title = "Do the conclusions depend on how missing values were handled?",
       x = "Coefficient", y = NULL, colour = NULL)
```

## Sensitivity 2 --- multiple imputation with Rubin's rules

```{r sens-mice, eval=PKG$mice && CFG$run_mice, cache=TRUE}
mice_vars <- c("life_expectancy", IMPUTE_VARS, "status", "year_c")
mice_df   <- as.data.frame(life[, intersect(mice_vars, names(life))])

imp_obj <- mice::mice(mice_df, m = CFG$mice_m, maxit = CFG$mice_maxit,
                      method = "pmm", seed = CFG$seed, printFlag = FALSE)

fits <- lapply(seq_len(CFG$mice_m), function(i) {
  di <- mice::complete(imp_obj, i) |>
    mutate(log_gdp = log1p_safe(gdp), log_population = log1p_safe(population),
           log_measles = log1p_safe(measles), log_hiv = log1p_safe(hiv_aids))
  lm(form_red, data = di)
})

pool_rubin <- function(fits) {
  m   <- length(fits)
  cs  <- lapply(fits, function(f) summary(f)$coefficients)
  tms <- rownames(cs[[1]])
  Q <- vapply(cs, function(e) e[tms, 1], numeric(length(tms)))
  U <- vapply(cs, function(e) e[tms, 2]^2, numeric(length(tms)))
  Qbar <- rowMeans(Q); Ubar <- rowMeans(U)
  B  <- apply(Q, 1, stats::var)
  Tv <- Ubar + (1 + 1 / m) * B
  se <- sqrt(Tv)
  lambda <- (1 + 1 / m) * B / Tv                  # fraction of missing information
  df <- ifelse(lambda > 0, (m - 1) / lambda^2, Inf)
  tibble(term = tms, estimate = Qbar, std.error = se,
         statistic = Qbar / se,
         p.value = 2 * stats::pt(-abs(Qbar / se), df),
         fmi = lambda)
}

pooled <- pool_rubin(fits)

pooled |>
  left_join(tidy_fit(m_red_imp, conf = FALSE) |>
              select(term, single_estimate = estimate, single_se = std.error),
            by = "term") |>
  mutate(se_inflation = std.error / single_se) |>
  select(term, estimate, single_estimate, std.error, single_se, se_inflation, fmi, p.value) |>
  arrange(desc(fmi)) |>
  kable(digits = 4,
        caption = "Pooled multiple-imputation estimates vs single imputation")
```



```{r sens-mice-2l, eval=FALSE}
# Research-quality version: country enters the imputation model as a cluster.
pred <- mice::make.predictorMatrix(mice_df2)          # mice_df2 includes a country id
pred[, "country_id"] <- -2                            # -2 marks the class variable
meth <- mice::make.method(mice_df2)
meth[IMPUTE_VARS] <- "2l.pmm"                         # or "2l.norm" / "2l.lmer"
imp2l <- mice::mice(mice_df2, method = meth, predictorMatrix = pred,
                    m = 5, maxit = 10, seed = CFG$seed)
```

## Sensitivity 3 --- influential observations

```{r sens-influence}
cook_cut  <- 4 / nrow(dat)
keep_rows <- which(stats::cooks.distance(m_red) <= cook_cut)
m_red_noinf <- lm(form_red, data = dat[keep_rows, ])

cat("Observations excluded:", nrow(dat) - length(keep_rows),
    sprintf("(%.1f%%)\n", 100 * (nrow(dat) - length(keep_rows)) / nrow(dat)))

infl_cmp <- tidy_fit(m_red, conf = FALSE) |>
  select(term, with_all = estimate) |>
  left_join(tidy_fit(m_red_noinf, conf = FALSE) |>
              select(term, without_influential = estimate), by = "term") |>
  mutate(change_pct = 100 * (without_influential - with_all) / abs(with_all),
         sign_flip = sign(with_all) != sign(without_influential))

infl_cmp |> arrange(desc(abs(change_pct))) |>
  kable(digits = 3,
        caption = "Effect of removing high-influence observations on the coefficients")

cat(sprintf("Sign flips: %d | largest relative change: %.1f%% (%s)\n",
            sum(infl_cmp$sign_flip, na.rm = TRUE),
            max(abs(infl_cmp$change_pct), na.rm = TRUE),
            infl_cmp$term[which.max(abs(infl_cmp$change_pct))]))
```

## Sensitivity 4 --- predictor set

```{r sens-predset, cache=TRUE}
form_lasso <- if (PKG$glmnet && exists("cv_list") && exists("extract_coefs")) {
  sel <- extract_coefs(cv_list[["alpha_1"]], "Lasso") |>
    filter(estimate != 0) |> pull(term)
  sel <- intersect(make.names(sel), make.names(c(PRED_FULL, "year_c")))
  if (length(sel) < 2) form_red else
    stats::reformulate(c(sel, "status"), response = "life_expectancy")
} else form_red

sens_specs <- list(
  "Full predictor set"  = function(P) stats::predict(lm(form_full,  data = P$tr), newdata = P$te),
  "VIF-pruned set"      = function(P) stats::predict(lm(form_red,   data = P$tr), newdata = P$te),
  "Lasso-selected set"  = function(P) stats::predict(lm(form_lasso, data = P$tr), newdata = P$te)
)

summarise_cv(run_cv(foldid, sens_specs, data = panel)) |>
  kable(digits = 3, caption = "Grouped-CV error under three predictor sets")
```

## Sensitivity 5 --- how much did the leakage actually cost?

```{r sens-leakage, cache=TRUE}
# WRONG ON PURPOSE: preprocessing already applied to the whole panel, then split.
run_cv_leaky <- function(folds, specs, data = dat_imp) {
  out <- list()
  for (f in sort(unique(folds))) {
    itr <- which(folds != f); ite <- which(folds == f)
    tr <- data[itr, ]; te <- data[ite, ]
    M  <- build_matrix(tr, te)
    P  <- list(tr = tr, te = te, Xtr = M$Xtr, Xte = M$Xte,
               ytr = tr$life_expectancy, yte = te$life_expectancy)
    for (nm in names(specs)) {
      pred <- tryCatch(specs[[nm]](P), error = function(e) rep(NA_real_, length(ite)))
      out[[length(out) + 1]] <- tibble(model = nm, fold = f, obs = P$yte,
                                       pred = as.numeric(pred))
    }
  }
  bind_rows(out)
}

leak_specs <- specs_grouped[intersect(c("Multiple regression", "Elastic net",
                                        "Random forest"), names(specs_grouped))]
leak_cmp <- bind_rows(
  summarise_cv(run_cv_leaky(foldid, leak_specs)) |>
    mutate(pipeline = "Impute full panel -> split (leaky)"),
  summarise_cv(run_cv(foldid, leak_specs, data = panel)) |>
    mutate(pipeline = "Split -> fit preprocessing on train (correct)")
) |>
  select(pipeline, model, RMSE_oof, R2_oof) |>
  arrange(model, pipeline)

kable(leak_cmp, digits = 4,
      caption = "Reported error under the leaky and the corrected pipelines")
```

```{r sens-leakage-gap}
leak_gap <- leak_cmp |>
  group_by(model) |>
  summarise(rmse_understated_by =
              RMSE_oof[grepl("correct", pipeline)] - RMSE_oof[grepl("leaky", pipeline)],
            .groups = "drop")
kable(leak_gap, digits = 4,
      caption = "How much the leaky pipeline understated the error (years of RMSE)")
```

## Verdict

```{r sens-verdict}
tibble(
  perturbation = c("Complete case vs imputed", "Single vs multiple imputation",
                   "Imputation model ignores country", "With vs without influential points",
                   "Predictor set", "Grouped CV vs temporal split",
                   "Leaky vs fold-wise preprocessing"),
  affects = c("Coefficient magnitudes (small), sample composition (large)",
              "Standard errors (inflated), point estimates (stable)",
              "Understates between-country variation - documented, not fixed",
              "Coefficient magnitudes (small)",
              "Predictive error (small), coefficient interpretation (large)",
              "Model ranking (moderate) - the two schemes answer different questions",
              "Reported error (measured above); the correct pipeline is the one used")
) |>
  kable(caption = "Summary of the sensitivity analyses")
```
